---
title: "R Notebook"
output: 
  html_notebook: 
    highlight: tango
---
# Lab 1 - Dowloading R and RStudio - No new code associated with lab
# Lab 2 - Working with Data in R
```{r Working with Data in R, echo=FALSE}
# This is a comment
iris <- iris
View(iris)

View(x = iris)

#Extract a single column
iris$Sepal.Length

#Save column as a separate variable
sep_len <- iris$Sepal.Length

#Run a function on a single column in df
mean(x = iris$Sepal.Length)

#Run a function on the 'sep_len' object
mean(x = sep_len) #Output should be identical to output from line 14

install.packages("tidyverse")
library(tidyverse)

#Use filter function in dplyr
# Base R command: iris[iris$Sepal.Length <5, ]
filter(iris, Sepal.Length < 5)

#Save the filtered output as a new variable
filtered_iris <- filter(iris, Sepal.Length < 5)

#Filter the iris dataset
# Base R command: iris[iris$Species == "setosa", ]
filter(iris, Species == "setosa")

#Filter iris dataset for the 'setosa' species and a 'Sepal.Length' less than 5
# Base R command: iris[iris$Species == "setosa", ]
filter(iris, Species == "setosa" & Sepal.Length < 5)


#Use the mutate() function
# Base R command: iris$new_data1 <- iris$Sepal.Length*2
mutate(iris, new_data1 = Sepal.Length *2)

#Add new columns 
mutate(iris, new_data1 = Sepal.Length *2,
       new_data2 = Sepal.Length/Sepal.Width) 


#Use the select() function
# Base R command: iris$Species
select(iris, Species) #selects the Species column and maintains the data structure


#Calculate the mean of Sepal.Length.
# Base R command: mean(iris$Sepal.Length)
summarise(iris, mean = mean(Sepal.Length))

#Calculate the mean, standard deviation and variance of Sepal.Length simultaneously
# Base R command: mean(iris$Sepal.Length); sd(iris$Sepal.Length); var(iris$Sepal.Length)
summarise(iris, mean = mean(Sepal.Length),
          sd = sd(Sepal.Length),
          variance = var(Sepal.Length)) #calculates the mean, standard deviation and variance of Sepal.Length and assigns them to a specified name for easy identification


#Use the arrange() function
# arrange(iris, Sepal.Length). This does the same thing as below
iris %>% arrange(Sepal.Length) # NOTE: ascending is default

#Arrange data in descending order
iris %>% arrange(desc(Sepal.Length)) # does the same thing as above but in descending order of Sepal.Length.

```

# Lab 3 - Plotting Sample Data Using ggplot

# Lab 4 - Plotting Real Data Using ggplot

# Lab 5 - Central Limit Theorem

# Lab 6 - Normality and Sample Properties 1

# Lab 7 - Normality and Sample Properties 2
```{r Normality and Sample Properties 2, echo=FALSE}
# Setup and Import ----
library(car)
library(pastecs)
library(tidyverse)
library(ggplot2)

# setwd("~/Module_4_Assumptions_Correlation/")
# Import data from the Allen Brain Institue 
load("ABI.RData")
source("functions.R")

summary(ABI)
# Histogram
ggplot(data = ABI, mapping = aes(x = ptau)) + 
  geom_histogram(mapping = aes(y = ..density..), bins = 30, fill = 'gray', color = 'black') + 
  stat_function(fun = dnorm, args = list(mean = mean(ABI$ptau, na.rm = TRUE), sd = sd(ABI$ptau, na.rm = TRUE)), color = 'red')

# Q-Q plot
ggplot(data = ABI, mapping = aes(sample = ptau)) + 
  geom_qq() + 
  geom_qq_line()

# Boxplot
ggplot(data = ABI, mapping = aes(x = ptau)) + 
  geom_boxplot() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())

# Histogram
ggplot(data = ABI, mapping = aes(x = ptau)) + 
  geom_histogram(bins = 30, fill = 'gray', color = 'black') +
  facet_wrap(~ struct, scales = "free")

# Q-Q plot
ggplot(data = ABI, mapping = aes(sample = ptau)) + 
  geom_qq() + 
  geom_qq_line() +
  facet_wrap(~ struct, scales = "free")

# Boxplot
ggplot(data = ABI, mapping = aes(x = ptau)) + 
  geom_boxplot() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  facet_wrap(~ struct, scales = "free")


# Test Normality -----
stat.desc.clean(dataset = ABI, variable = ptau)
stat.desc.clean(dataset = ABI, variable = ptau, struct)

# Transform to Normal distribution
ABI2 <- ABI %>% 
  mutate(ptau_sqrt = sqrt(ptau),
         ptau_log = log(ptau)) %>% 
  select(ID, struct, ptau, ptau_sqrt, ptau_log) %>% 
  pivot_longer(cols = ptau:ptau_log,
               names_to = "Measurement",
               values_to = "Concentration")

# Histogram
ggplot(data = ABI2, mapping = aes(x = Concentration)) + 
  geom_histogram(bins = 30, fill = 'gray', color = 'black') +
  facet_wrap(~ struct * Measurement, scales = "free")

# Q-Q plot
ggplot(data = ABI2, mapping = aes(sample = Concentration)) + 
  geom_qq() + 
  geom_qq_line() +
  facet_wrap(~ struct * Measurement, scales = "free")

# Boxplot
ggplot(data = ABI2, mapping = aes(x = Concentration)) + 
  geom_boxplot() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  facet_wrap(~ struct * Measurement, scales = "free")

stat.desc.clean(dataset = ABI2, variable = Concentration, struct, Measurement)

# Homogeneity of Variance
leveneTest(ptau ~ struct, data = ABI)


mip_ABI <- ABI %>% 
  mutate(mip_sqrt = sqrt(mip_1a),
         mip_log = log(mip_1a)) %>% 
  select(ID, struct, mip_1a, mip_sqrt, mip_log) %>% 
  pivot_longer(cols = mip_1a:mip_log,
               names_to = "Measurement",
               values_to = "Concentration")

# Histogram
ggplot(data = mip_ABI, mapping = aes(x = Concentration)) + 
  geom_histogram(bins = 30, fill = 'gray', color = 'black') +
  facet_wrap(~ struct * Measurement, scales = "free")

# Q-Q plot
ggplot(data = mip_ABI, mapping = aes(sample = Concentration)) + 
  geom_qq() + 
  geom_qq_line() +
  facet_wrap(~ struct * Measurement, scales = "free")

# Boxplot
ggplot(data = mip_ABI, mapping = aes(x = Concentration)) + 
  geom_boxplot() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  facet_wrap(~ struct * Measurement, scales = "free")

stat.desc.clean(dataset = mip_ABI, variable = Concentration, struct, Measurement)

# Homogeneity of Variance
leveneTest(mip_sqrt ~ struct, data = mip_ABI)

```


# Lab 8 - Correlation
```{r Correlation, echo=FALSE}
# Objective: Perform basic correlation analysis with base R cor function and review Pearson's, Spearman's, and Kendall's correlations.

```


# Lab 9 - Linear Regression 1
```{r Lab 9. Loading pkgs and data}
library(tidyverse)
theme_set(theme_bw())
setwd("./Module_5_Linear_Regression/")
load("Module_5_Linear_Regression/framingham.RData")
source("Module_3_CLT-and-Normality/functions.R")
```

```{r Lab 9. Getting to know your data}
head(fhs)
summary(fhs)

# Transforming the data to make it easy to plot with facets
fhs2 <- fhs %>% 
  select(-gender) %>% 
  pivot_longer(cols = age:glucose,
               names_to = "Metric",
               values_to = "Value")

# Distribution of the variables
ggplot(data = fhs2, mapping = aes(x = Value)) + 
  geom_histogram(bins = 30, fill = 'gray', color = 'black') +
  facet_wrap(~ Metric, scales = "free")

# Relationship between the variables
fhs3 <- fhs %>% 
  select(-gender) %>% 
  pivot_longer(cols = c(age, glucose),
               names_to = "Metric",
               values_to = "Value")

ggplot(data = fhs3, mapping = aes(x = Value, y = sysBP)) +
  geom_point() +
  facet_wrap(~ Metric, scales = "free")
```

```{r Lab 9. Assumptions}
# We will review this after the next class
```

```{r Lab 9. Simple linear regression}
# Formula outcome (dependent) ~ predictor (independent), y ~ x

# Model systolic BP as a function of age
sysBPvsAge <- lm(formula = sysBP ~ age, data = fhs)

summary(sysBPvsAge)

# Subset the coefficients table by subsetting the model summary
summary(sysBPvsAge)$coefficients
# sysBPvsAge$coefficients is equivalent to sysBPvsAge[["coefficients"]] and stats::coef(sysBPvsAge)
```


# Lab 13. ANCOVA
```{r Lab 13. ANCOVA}
library(car)
library(ez)
library(rstatix)
library(pastecs)
library(QuantPsyc)
library(tidyverse)
setwd("./Module_7_ANOVA_GLM")

load("/Users/claireps/GBSC_731_Intro-to-Biostats/Module_7_ANOVA_GLM/ANCOVA_Data.RData")
source("/Users/claireps/GBSC_731_Intro-to-Biostats/Module_3_CLT-and-Normality/functions.R")
```

## Getting to Know Your Data
```{r}
summary(mouse.mass) 

ggplot(data = mouse.mass, mapping = aes(x = group, y = bodyfat)) + 
  geom_bar(mapping = aes(fill = group), stat = "summary", fun = "mean", show.legend = FALSE) + 
  geom_errorbar(stat = "summary", fun.data = "mean_se", width = 0.3) + 
  geom_jitter(width = 0.3) + 
  labs(title = "Mean body fat in mice derived from different environments")

ggplot(data = mouse.mass, mapping = aes(x = group, y = leanmass)) + 
  geom_bar(mapping = aes(fill = group), stat = "summary", fun = "mean", show.legend = FALSE) + 
  geom_errorbar(stat = "summary", fun.data = "mean_se", width = 0.3) + 
  geom_jitter(width = 0.3) + 
  labs(title = "Mean lean mass in mice derived from different environments")
```

## Test Assumptions
```{r}
## Homogeneity of Regression Slopes and Linearity ----
ggplot(data = mouse.mass, aes(x = leanmass, y = bodyfat, color = group)) +
  geom_point() + 
  geom_smooth(method = 'lm', se = FALSE)


mouse.mass %>% 
  filter(group == 'wild') %>% 
  lm(bodyfat ~ leanmass, data = .) %>% 
  lm.beta()

mouse.mass %>% 
  filter(group == 'wild-der') %>% 
  lm(bodyfat ~ leanmass, data = .) %>% 
  lm.beta()

mouse.mass %>% 
  filter(group == 'lab') %>% 
  lm(bodyfat ~ leanmass, data = .) %>% 
  lm.beta()
```

## Running ANOVA
### Regular ANOVA
```{r}
mod <- ezANOVA(data = mouse.mass,
               dv = bodyfat,
               between = group,
               wid = idnum,
               type = 3,
               return_aov = TRUE)

mod

# Extracting residuals
residuals <- tibble(group = mouse.mass$group,
                    resid = resid(mod$aov))

# Normality within groups
ggplot(data = residuals, mapping = aes(x = resid)) + 
  geom_histogram(bins = 20, fill = 'gray', color = 'black') +
  facet_wrap(~ group, scales = "free")

ggplot(data = residuals, mapping = aes(sample = resid)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ group, scales = "free")

ggplot(data = residuals, mapping = aes(x = resid)) +
  geom_boxplot() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  facet_wrap(~ group, scales = "free")

stat.desc.clean(dataset = residuals, variable = resid, group)
```

### ANCOVA
```{r}
mod <- anova_test(data = mouse.mass,
                  dv = bodyfat,
                  between = group,
                  covariate = leanmass,
                  wid = idnum,
                  type = 3)

mod

# Extracting residuals
residuals <- tibble(group = mouse.mass$group,
                    resid = resid(attributes(mod)$args$model))

# Normality within groups
ggplot(data = residuals, mapping = aes(x = resid)) + 
  geom_histogram(bins = 20, fill = 'gray', color = 'black') +
  facet_wrap(~ group, scales = "free")

ggplot(data = residuals, mapping = aes(sample = resid)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~ group, scales = "free")

ggplot(data = residuals, mapping = aes(x = resid)) +
  geom_boxplot() +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  facet_wrap(~ group, scales = "free")

stat.desc.clean(dataset = residuals, variable = resid, group)

# Homogeneity of variance - wasn't tested by anova_test
leveneTest(resid ~ group, data = residuals)
```

